---
title: "UN_Data_Exploration_maria"
output: html_notebook
---

```{r}
library(tidyverse)
```


1. Using the read_csv() function, read the GDP dataset into your notebook as a tibble called gdp_df. After reading it in, inspect the first 10 rows and then inspect the last 10 rows.

```{r}
gdp_df <- read_csv("../data/gdp_per_capita.csv")
```

```{r}
gdp_df %>% 
  slice_head(n=10)
# also:
# slice(1:10)
```

```{r}
gdp_df %>% 
  slice_tail(n=10)
```

2. Drop the 'Value Footnotes' column, and rename the remaining columns to 'Country', 'Year', and 'GDP_Per_Capita'.

```{r}
gdp_df <- gdp_df |> 
  select(-`Value Footnotes`) |> 
  rename(Country = `Country or Area`, GDP_Per_Capita = Value)
```

3. Which years are represented in this dataset? How many observations are there per year? Make a plot to view the number of observations per year.

```{r}
year_count <- gdp_df |> 
  count(Year) |> 
  rename(count = n) 
print(year_count |> summarise(max(Year), min(Year)))
```
```{r}
ggplot(year_count, aes(x = Year, y = count)) +
  geom_col()
```

4. How many countries are represented in this dataset? Which countries have the fewest observations?

```{r}
country_count <- gdp_df |> 
  count(Country, sort=TRUE) |> 
  rename(count = n)
```

```{r}
sprintf('%s countries are represented in the dataset.', n_distinct(country_count))
```
```{r}
country_count |> 
  slice_min(count)
```
5. Create a new tibble by subsetting gdp_df to just the year 2021. Call this new tibble gdp_2021.

```{r}
gdp_2021 <- gdp_df |> 
  filter(Year == 2021)
```

6. Use summary() to find the summary statistics for GDP per capita in 2021.

```{r}
summary(gdp_2021)
```
7. Create a histogram of GDP Per Capita numbers for 2021 (you may wish to adjust the number of bins for your histogram).

```{r}
ggplot(gdp_2021, aes(x = GDP_Per_Capita)) +
  geom_histogram(bins=20)
```
8. Find the top 5 counties and bottom 5 countries by GDP per capita in 2021.

```{r}
slice_max(gdp_2021, GDP_Per_Capita, n=5)
slice_min(gdp_2021, GDP_Per_Capita, n=5)
```
9. Now, return to the full dataset, gdp_df. Pivot the data for 1990 and 2021 (using pivot_wider() function) so that each row corresponds to a country, each column corresponds to a year, and the values in the table give the GDP_Per_Capita amount. Drop any rows that are missing values for either 1990 or 2021. Save the result to a tibble named gdp_pivoted.

```{r}
gdp_pivoted <- gdp_df |> 
  filter(Year %in% c(1990, 2021)) |>
  pivot_wider(id_cols=Country, names_from=Year, values_from=GDP_Per_Capita) |> 
  drop_na()
```

10. Create a new column in gdp_pivoted named Percent_Change. This column should contain the percent change in GDP_Per_Capita from 1990 to 2021. Hint: Percent change is calculated as 100*(New Value - Old Value) / Old Value.

```{r}
gdp_pivoted <- gdp_pivoted |> 
  mutate(Percent_Change = 100*(`2021`-`1990`)/`1990`)
```

11. How many countries experienced a negative percent change in GDP per capita from 1990 to 2021?

```{r}
gdp_pivoted |>
  filter(Percent_Change < 0) |>
  n_distinct()
```
12. Which countries had the two highest % change values in GDP per capita? Create a line plot showing these countries' GDP per capita for all years for which you have data. Put both line charts on the same plot.

```{r}
slice_max(gdp_pivoted, Percent_Change, n=2)
```
```{r}
top2_list <- slice_max(gdp_pivoted, Percent_Change, n=2) |> 
  select(Country) |> 
  deframe()
top2 <- gdp_df |> 
  filter(Country %in% c(top2_list))
ggplot(top2, aes(x=Year, y=GDP_Per_Capita, colour=Country)) +
  geom_line()
```
13. Read in continents.csv contained in the data folder into a new tibble called continents. We will be using this tibble to add a new column to our dataset.

