---
title: "UN_Data_Exploration_maria"
output: html_notebook
---

```{r}
library(tidyverse)
```


1. Using the read_csv() function, read the GDP dataset into your notebook as a tibble called gdp_df. After reading it in, inspect the first 10 rows and then inspect the last 10 rows.

```{r}
gdp_df <- read_csv("../data/gdp_per_capita.csv")
```

```{r}
gdp_df %>% 
  slice_head(n=10)
# also:
# slice(1:10)
```

```{r}
gdp_df %>% 
  slice_tail(n=10)
```

2. Drop the 'Value Footnotes' column, and rename the remaining columns to 'Country', 'Year', and 'GDP_Per_Capita'.

```{r}
gdp_df <- gdp_df |> 
  select(-`Value Footnotes`) |> 
  rename(Country = `Country or Area`, GDP_Per_Capita = Value)
```

3. Which years are represented in this dataset? How many observations are there per year? Make a plot to view the number of observations per year.

```{r}
year_count <- gdp_df |> 
  count(Year) |> 
  rename(count = n)
```

```{r}
glimpse(year_count)
```


```{r}
ggplot(year_count, aes(x = Year, y = count)) +
  geom_col()
```

4. How many countries are represented in this dataset? Which countries have the fewest observations?

```{r}
country_count <- gdp_df |> 
  count(Country, sort=TRUE) |> 
  rename(count = n)
```

```{r}
sprintf('%s countries are represented in the dataset.', n_distinct(country_count))
```

